---
title: "Description and usage of MetClassNetR: MTBLS1586"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of Description and usage of MetClassNetR}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{Spectra}
    %\VignetteDepends{Spectra,BiocStyle}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

```{r, echo = FALSE, message = FALSE}
library(BiocStyle)
```

# Introduction

XXX (Description of MTBLS1586)

# Setup

## Load required libries

<code>MetClassNetR</code> integrates functions from different libraries.
Install current version of <code>MetClassNetR</code> and additional 
<code>MetNet</code> functions from developmental branches. After 
Installation re-load R and load required libraries.

```{r libraries, message=FALSE, warning=FALSE}
# installation of additional packages
#devtools::install_github("tnaake/MetNet") 
# note: Re-load R studio after installation
devtools::install_github("MetClassNet/MetClassNetR", 
                         ref = "devel_liesa") 

# load required libraries
library(MetClassNetR)
library(Spectra)
library(MsCoreUtils)
library(MsBackendMgf)
library(easyGgplot2)
library(QFeatures)
library(tidyverse)
#library(nontarget)
# library(visNetwork) # for visNetwork() and friends
# library(networkD3)  # for saveNetwork()
```
## Parallelization 
Parallelization can be done to speed up calculations. 
```{r}
register(bpstart(SnowParam(4, progressbar = TRUE)))
```

## Reading of MS1 data

MS1 data is read from Metabolights MAF or mzTab-M files into a 
<code>QFeatures</code> object. Information about the mass-to-charge 
(mz) ratio and retention times (RT) were extracted from the MAF/ mzTab-M 
file and stored into the <code>QFeatures</code>-object. The data from 
positive and negative ionization mode are stored in two different 
<code>QFeatures</code>-objects. 

```{r ms1, message=FALSE}
# Metabolights MAF files
pos_maf <- system.file("extdata/MTBLS1586/m_MTBLS1586_LC-MS_positive_reverse-phase_metabolite_profiling_v2_maf.tsv",
                       package = "MetClassNetR")
neg_maf <- system.file("extdata/MTBLS1586/m_MTBLS1586_LC-MS_positive_reverse-phase_metabolite_profiling_v2_maf.tsv",
                       package = "MetClassNetR")

# create QFeature objects
mtbls1586_qf_pos <- readMaf(file(pos_maf), ecol = 23, name = "mtbls1586_pos")
mtbls1586_qf_neg <- readMaf(file(neg_maf), ecol = 23, name = "mtbls1586_neg")
```


## Reading of MS2 data

MS2 data is read from a .mgf or .msp file into a <code>Spectra</code> object.

```{r ms2, message=FALSE, warning=FALSE}
# get MS2 data in .mgf files
pos_mgf <- system.file("extdata/MTBLS1586/ms2_MTBLS1586_LC-MS_positive_reverse-phase_metabolite_profiling.mgf",
                       package = "MetClassNetR")
neg_mgf <- system.file("extdata/MTBLS1586/ms2_MTBLS1586_LC-MS_negative_reverse-phase_metabolite_profiling.mgf",
                       package = "MetClassNetR")

# load MS2 data 
ms2_spectra_pos <- Spectra::Spectra(pos_mgf,
                           source = MsBackendMgf(),
                           backend = MsBackendDataFrame())

ms2_spectra_neg <- Spectra::Spectra(neg_mgf,
                           source = MsBackendMgf(),
                           backend = MsBackendDataFrame())
```

## Sanity checks

check data (namespace of ids etc...)

```{r}
# check minimal requirements
checkQFeatures(mtbls1586_qf_neg)
checkQFeatures(mtbls1586_qf_pos)
checkSpectra(ms2_spectra_neg)
checkSpectra(ms2_spectra_pos)
```

<code>checkIDNamespace</code> sanity check will fail if Ids between 
<code>QFeatures</code> and <code>Spectra</code> are not matching.Therefore, 
MS2 spectra can be filled with empty spectra to have a complete list of spectra 
for all the features.

```{r}
# # fill spectra
# ms2_spectra_neg <- fillSpectra(mtbls1586_qf_neg, ms2_spectra_neg)
# ms2_spectra_pos <- fillSpectra(mtbls1586_qf_pos, ms2_spectra_pos)
# 
# # check namespace 
# checkIdNamespace(mtbls1586_qf_neg, ms2_spectra_neg)
# checkIdNamespace(mtbls1586_qf_pos, ms2_spectra_pos)
```


## Reading of transformation file

The mass-difference network is calculated using the <code>MetNet</code> 
package. Here, all calculated mass-differences between all pairs of features 
are compared to a transformation-list, containing different biochemical 
reactions and their corresponding mass difference. Required columns are 
"group" and "mass". Additional columns as "formula" can be added.
The transformation-list needs to be provided by the user. An example 
can be found in the system files. 

```{r transformation}
transformations_file <- system.file("extdata/MTBLS1586/transformations_MTBLS1586.csv", 
                                    package = "MetClassNetR")

transformations <- read_csv(transformations_file, col_names = TRUE)

transformations[1:5,]
```


## Quality check of input

ZZZ

## Setup for networks

The <code>master</code>-list will contain all types of experimental 
networks that have been created in this workflow. 

```{r}
master <- list()
```

# Analysis

## Mass difference Network
The mass-difference network is calculated using the <code>MetNet</code> 
package. All mass-differences between all pairs of features are calculated 
and are compared to the <code>transformation</code>-list. This 
<code>transformation</code>-list contains different biochemical reactions 
and their corresponding mass difference. Only if the mass-difference 
between two features matches to a mass-difference in the transformation 
file, within a certain <code>ppm</code>-range, it will be displayed. 
Different to the original <code>MetNet</code>-functions, not only the 
name of the mass-difference but also the value itself will be displayed 
as output adjacency matrix.
The adjacency matrix will be than transformed to a <code>data.frame</code>. 

```{r mass_difference}
# Create mass difference network
massdiff <- qfeat_structural(x = mtbls1586_qf_pos,
                             assay_name = "mtbls1586_pos",
                             transformation = transformations,
#                             var = c("Name", "Formula difference", "mass"),
                             var = c("name", "formula", "mass"),
                             ppm = 10)

# Create data.frame of adjacency matrix
massdiff_df <- as.data.frame(massdiff) |> 
  filter(binary == 1)

massdiff_df[5:10,]
```

A summary of the current data-set on the present mass-difference 
distribution is provided using the <code>summary_mz</code>-function. 
This function stores the summary as data.frame, which is also plotted. 
Depending on the size of the data, it might be useful to <code>filter</code> 
the number of determined mass-differences, e.g. by 1000 counts.

```{r}
# Summary & filter for counts e.g. higher 1000
#sum_mz_f <- mz_summary(massdiff, filter = 1000, var = "group")

# some visualizations
#mz_vis(sum_mz_f, var = "group")

# append mass difference network
master <- c(master, massdiff)

```


## Add annotatation 
A <code>QFeatures</code>-object stores various information on LC-MS/MS 
data. Also, annotations and possible database identifiers are present 
in a <code>QFeatures</code>-object. In order to add the annotations to 
the adjacency-list, <code>qfeat_annotation</code> is applied. It extracts 
the annotations, if any present, and adds them to the desired 
adjacency-list, for example the list from the mass-difference network. 
In order to check if the annotations match the mass-difference, a filter 
is applied so that only features remains that have annotations in both 
features. 

```{r manual_annotation}
# Add annotations to mass-difference adjacency matrix
rowData(massdiff) <- as.data.frame(rowData(mtbls1586_qf_pos[["mtbls1586_pos"]]))

# display annotation for the feature "Cluster_3201"
rowData(massdiff)["Cluster_3201", ]

massdiff_anno <- as.data.frame(rowData(massdiff))

# Filter for present annotations
anno_filter <- filter(massdiff_anno, 
                      massdiff_anno$metabolite_identification != "")
anno_filter[25:30,]

```


## Correlation network

Correlation networks will be calculated using functionality from 
the <code>MetNet</code> package. <code>qfeat_statistical</code> 
uses the <code>QFeatures</code>-Object as input and extracts the 
feature intensity in order to calculate correlations from the 
<code>statistical</code>-function from <code>MetNet</code>. 
<code>MetNet</code> implemented various models that can be applied 
to calculate correlations, e.g. pearson, spearman, partial-pearson, 
and many more which might also be selected. After generating the 
correlation adjacency-matrix, it will also be transformed to a 
<code>data.frame</code> 

```{r correlation, warning=FALSE, error=FALSE}
# Correlation network using partial pearson and GGM correlation
corr <- qfeat_statistical(x = mtbls1586_qf_pos, 
                                assay_name = "mtbls1586_pos", 
                                model = c("pearson_partial", "ggm"),
                                na.omit = TRUE,
                                p.adjust = "BH")

# Create data.frame from correlation adjacency matrix 
corr_df <- as.data.frame(corr)
corr_df[25:30,]

master <- c(master, corr)

## plots to view the distribution of correlations 
corr_fl <- data.frame(coef = c(corr_df$pearson_partial_coef,
                              corr_df$ggm_coef),
                     group = c(rep("pearson_partial", nrow(corr_df)),
                               rep("ggm", nrow(corr_df)))
                     )

plot <- ggplot2.histogram(data=corr_fl, xName='coef',
                  groupName='group', legendPosition="right",
                  alpha=0.5, binwidth=0.01,
                  brewerPalette="Paired",
                  addMeanLine=TRUE,  meanLineSize=1)

plot_custom <-ggplot2.customize(plot, xtitle="Correlation coefficient", 
                                ytitle="Count",
                        showLegend=TRUE, axisLine=c(0.5, "solid", "black"),
                        addDensity=TRUE, removePanelBorder=TRUE, 
                        backgroundColor="white", 
                        mainTitle="Correlation coefficient")   

plot_custom


```


## Homologue Series

ABC


```{r homolseries}
# homol <- qfeat_homol(x = mtbls1586_qf_pos,
#                              assay_name = "mtbls1586_pos",
#                       elements=c("C","H","O"), use_C=TRUE,
#                       minmz=5, 	maxmz=50,
#                       minrt=5,  maxrt=60,
#                       ppm=FALSE,
#                       mztol= 0.005,  rttol=5,
#                       minlength=4,
#                       mzfilter=FALSE,
#                       spar=.45, 	R2=.98,
#                       plotit=FALSE)
```


## MS2 similarity network

Spectral similarity network is created using positive MS2 spectra.

```{r ms2_network}
tolerance = 0.005
ppm = 0

# filter ms2 spectra 20 % of intensity
filter <- TRUE
my_filter <- function(x) {
  x > max(x, na.rm = TRUE) / 10
}


ms2_spectra_pos <- filterIntensity(ms2_spectra_pos, my_filter)

## to speed up calculations: only use 50 spectra for spectral similarity
spect <- spec_molNetwork(ms2_spectra_pos[1:50],
                             MAPFUN = joinPeaksGnps,
                             methods = "gnps",
                             tolerance = tolerance,
                             ppm = ppm, 
                             type = "inner")


spect_adj <- addSpectralSimilarity(am_structural = massdiff, 
                                   ms2_similarity = spect)

# Create data.frame of adjacency matrix
spect_df <- as.data.frame(spect_adj) |> 
  filter(!is.na(gnps))

master <- c(master, spect_adj)
```


# Export networks to gml format 

Created networks are exported and saved as .gml file in the current 
working-directory. Feature Ids will be stored as additional name in 
node attributes. All the other information from the adjacency-lists 
will be stores as edge-attributes. It might also be defined which 
columns of the adjacency-list should be stored, by using the 
<code>select</code> option in the <code>exportNet2gml</code>-function.

```{r export2gml, eval=FALSE}
## Mass-difference network to gml
# note: only first 10 features are exported
exportNet2gml(x = massdiff_df[1:10,], file = "mzdiff")

## Correlation network to gml
# note: only first 10 features are exported
exportNet2gml(x = pears_df[1:10,], file = "pears")

## Spectral similarity network2gml
# note: only first 10 features are exported
exportNet2gml(x = spect_df[1:10,], file = "similarity", 
              select = c("gnps"))

```

Additionally, attributes may be exported as .gml attribute-file using 
<code>exportAttributes2gml</code>. Edge attributes are prepared by 
merging the adjacency-lists from mass-difference and correlation-networks. 
Node attributes are generated by extraction additional information as 
annotation, mz, and RT values from the <code>QFeatures</code>-object. 

```{r, eval=FALSE}
# Create and export attribute file 
attributes <- merge(corr_df, massdiff_df, by = c("Row", "Col"))

# Prepare annotation file 
feat_anno <- as.data.frame(rowData(mtbls1586_qf_pos[["mtbls1586_pos"]]))
feat_anno <- feat_anno[,c("metabolite_identification", "mz", "rtime")]
colnames(feat_anno) <- c("Annotation", "mz", "RT")

  
# note: only first 10 features are exported
exportAttributes2gml(attributes[1:10,], file = "attributes_mtbls1586_pos", 
                     anno = feat_anno)
```

Export of homologous series
```{r graph_export}
# 
# el <- as_tibble(homol[["Peaks in homologue series"]]) |>
#     filter(`to ID` != "0") |> select (c("peak ID", "to ID",
#                                          "m/z increment", "RT increment"))
# 
# g <- el |> select (c("peak ID", "to ID")) |> as.matrix |> graph_from_edgelist 
# 
# ## Write to File
# write_graph(g, "/tmp/g.gml", "gml")

```




