---
title: "Data preparation for neo4j"
author: "Sarah Scharfenberg"
date: "4 8 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE}
#remove.packages("classyfireR", lib="~/R/x86_64-pc-linux-gnu-library/4.0")
#remotes::install_github('Flow191/classyfireR', ref="devel",build_vignettes = TRUE)

library(classyfireR)
conn <- classyfireR::open_cache("20210804_classyfireR.db")

library(MetClassNetR)
library(dplyr)
library(tibble)
```

## Read GSMN

```{r, message=FALSE}
file <- system.file("extdata/Example_Coverage/iHY3410_Compound-SBtab.tsv", package = "MetClassNetR")

network <- read.csv(file,
                    header = TRUE,
                    skip=1,
                    stringsAsFactors = FALSE,
                    sep="\t",
                    na.strings = "")

colnames(network) <- gsub(pattern = "X.",
                          replacement = "",
                          x = colnames(network))

```

### File as input
If a file is given as input each compound in the file will be processed. In case of GSMNs this could mean that the same compound occurring in different compartments will be classified multiple times.
```{r, message=FALSE}
file <- system.file("extdata/Example_Coverage/iHY3410_Compound-SBtab.tsv", package = "MetClassNetR")

network_tibble <- gsmn_to_ontology(file=file,ontology = "chemont", conn=conn)
```

### InChIKey vector as input
When the vector of InChIKeys is given it is possible to filter NA and duplicates before.
Although this would be the fastest way to get the classification it depends on the conversion tool that was used to generate the InChIKeys wether the classification is successfull or not. 
```{r, message=FALSE}
inchikeys <- network[,"Identifiers.inchikey"]
inchikeys <- inchikeys[!is.na(inchikeys)]
inchikeys <- unique(inchikeys)

network_tibble_2 <- gsmn_to_ontology(inchikeys = inchikeys, ontology = "chemont", conn=conn)
```

### SMILES vector as input
Using SMILES or InChIs as input the identifiers are converted using RCDK which is the same tool that is used by the webtool to get the InChIKey. The successrate is highest here.
```{r, message=FALSE}
smiles <- network[,"Notes.SMILES"]
smiles <- smiles[!is.na(smiles)]
smiles <- unique(smiles)

network_tibble_3 <- gsmn_to_ontology(smiles = smiles, ontology = "chemont", conn=conn)
```

### Comparison
Depending on the conversion tool the GSMN creators used it can happen that using SMILES will have more
compounds classified in the end.

```{r}
network_tibble %>%
  filter(!is.na(Identifiers.inchikey)) %>%
  select(Identifiers.inchikey, classified) %>%
  dplyr::distinct(Identifiers.inchikey, .keep_all=TRUE ) %>%
  filter(classified==TRUE) %>%
  select(classified) %>%
  count()

network_tibble_2 %>%
  filter(classified==TRUE) %>%
  select(classified) %>%
  count()

network_tibble_3 %>%
  filter(classified==TRUE) %>%
  select(classified) %>%
  count()

```




